# BIOL 4800/6220 Fall 2024 R Tutorial

## Practical Computing Fall 2024

```{r}
setwd("~/praccomp2024/")
```

### R Basics

#### Introduction

* R as a calculator

```{r}
1 + 100
```

```{r}
2^3
```

```{r}
7/2
```

```{r}
4-1
```

```{r}
2*3
```

```{r}
3+5*2
```

```{r}
(3+5)*2
```

```{r}
2e3
```

```{r}
2/10000
```

```{r}
log(1)
```

```{r}
exp(1)
```

* Boolean Conditionals

```{r}
1 == 1
```

```{r}
2 == 1
```

```{r}
2 > 1
```

```{r}
2 <= 1
```

```{r}
1 != 2
```

* Variables

```{r}
x <- 10
x
```

```{r}
x + x
```

```{r}
x - 5
```

```{r}
y <- x^x
```

```{r}
y
```

* Vectorization

```{r}
1:5
```

```{r}
2^(1:5)
```

```{r}
v <- 1:5
```

```{r}
2^v
```

```{r}
log(v)
```

* Environment

```{r}
ls()
```

```{r}
rm(v)
```

```{r}
ls()
```

#### Package Managment

```{r}
#installed.packages()
```

```{r}
#install.packages("vegan", dependencies = TRUE)
#library(vegan)
```

```{r}
#if (!require("BiocManager", quietly = TRUE))
#    install.packages("BiocManager")

#BiocManager::install("DESeq2")
```

```{r}
#library(DESeq2)
```

#### Getting Help

**Search the web**

```{r}
?plot
```

```{r}
help(plot)
```

```{r}
vignette("FAQ-vegan")
```

```{r}
citation("vegan")
```

#### Data and their formats

```{r}
cats <- data.frame(coat = c("calico", "black", "tabby"),
                   weight = c(2.1, 5, 3.2),
                   likes.string = c(TRUE, FALSE, TRUE))
```

```{r}
cats
```

```{r}
getwd()
```

```{r}
write.csv(cats, "cats.csv")
```

```{r}
cats <- read.csv("cats.csv", header=TRUE, row.names = 1)
```

```{r}
cats
```

```{r}
cats$weight
```

```{r}
cats[, 2]
```

```{r}
cats[, "weight"]
```

```{r}
cats[c(1,3),c(2:3,1)]
```

```{r}
(cats.log.weight <- log(cats$weight))
```

```{r}
cbind(cats,cats.log.weight)
```

```{r}
(cats2 <- cbind(cats,cats.log.weight))
```

```{r}
cats2
```

```{r}
paste("My cat is ", cats$coat, ", and it weighs ", cats$weight, "kg.", sep = "")
```

##### Data Types

```{r}
typeof(cats)
```

```{r}
class(cats)
```

```{r}
typeof(cats$coat)
typeof(cats$weight)
typeof(cats$likes.string)
```

```{r}
typeof(1i)
```

```{r}
likes.pop <- c(1,0,1,1,0)

typeof(likes.pop)
```

```{r}
as.logical(likes.pop)
```

```{r}
names <- c("Iggy", "Bitey", "Garfield")

cats$names <- names

cats
```

```{r}
names2 <- c(names, "Spike")
```

```{r}
names3 <- c("Fluffy", names)
```

```{r}
1:10
```

```{r}
(z <- seq(20, 150, by=0.5))
```

```{r}
length(z)
```

```{r}
length(cats)
```

```{r}
length(cats$coat)
```

```{r}
dim(cats)
```

```{r}
typeof(z)
```

```{r}
class(z)
```

```{r}
head(z, n=20)
```

```{r}
tail(z, n=40)
```

```{r}
sort(z, decreasing = TRUE)
```

```{r}
unique(z)
```

##### Data Frames

```{r}
cats
```

```{r}
cats$weight

typeof(cats$weight)
```

```{r}
str(cats$weight)
```

```{r}
coats <- c("tabby", "tortoiseshell", "tortoiseshell", "black", "tabby")
```

```{r}
(categories <- factor(coats))
```

```{r}
class(categories)
```

```{r}
str(categories)
```

##### Lists

```{r}
(list.example <- list(title="Numbers", numbers = seq(1,20,by=5), data=TRUE))
```

```{r}
list.example$title
```

```{r}
data.frame(list.example)
```

##### Matricies

```{r}
(matrix.example <- matrix(0, nrow = 3, ncol = 5, ))
```

```{r}
class(matrix.example)
```

```{r}
typeof(matrix.example)
```

```{r}
dim(matrix.example)
```

```{r}
nrow(matrix.example)
ncol(matrix.example)
```

```{r}
data.frame(matrix.example)
```

##### Subsetting

```{r}
(p <- c(2.3,6.9,4.0,23,1))
```

```{r}
names(p) <- c("a", "b", "c", "d", "e")
p
```

```{r}
typeof(p)
class(p)
```

```{r}
p[1]
```

```{r}
p[2:4]
```

```{r}
p[c(1,5)]
```

```{r}
p[c(1,1,1,3,5,5)]
```

```{r}
p[6]
```

```{r}
(q <- p[-3])
```

```{r}
p[-(2:4)]
```

```{r}
p[c("a", "e")]
```

```{r}
p[c(T,F,T,F,T)]
```

##### Factors

```{r}
(f <- factor(c("a", "b", "c", "d", "e")))
```

```{r}
f[f == "a"]
```

```{r}
f[1:3]
```

```{r}
(f2 <- factor(c("a", "a", "b", "c", "c")))
```

```{r}
f2[f2 == "a"]
```

```{r}
f2[f2 %in% c("a", "c")]
```

##### Matricies Revisted

```{r}
set.seed(2525)

m <- matrix(rnorm(6,4,0.3), nrow=6, ncol=4)
m
```

```{r}
hist(rnorm(10000,4,0.3))
```

```{r}
m[3:4, c(3,1)]
```

##### Lists Revisted

```{r}
xlist <- list(a="BIOL48006220", b=rnorm(16,75,10), data="Grade")
xlist
```

##### Data Frames Revisted

```{r}
(gp <- read.csv("data/forest_area_sq_km.csv"))
```

```{r}
head(gp, n=10)
```

```{r}
head(gp[4], n=3L)
```

```{r}
nrow(gp)
```

```{r}
head(gp[["country"]], 10L)
```
```{r}
gp[c(1,3), c(1, 5:10)]
```

```{r}
gp[ which(gp$X1989 <= 100 & gp$X2019 >= 100), ]
```

#### Conditionals and Flow

```{r}
n <- 1

if (n < 10) {
  print("n is less than 10.")
} else if (n > 10) {
  print("n is greater than 10.")
} else {
  print("n is equal to 10")
}
```

```{r}
for (n in seq(0,20)) {
  if (n < 10) {
    print("n is less than 10.")
}   else if (n > 10) {
    print("n is greater than 10.")
}   else {
    print("n is equal to 10")
}
}

```

```{r}
for (n in rnorm(20,10,3)) {
  if (n < 10) {
    print(paste0(n," is less than 10."))
}   else if (n > 10) {
    print(paste0(n," is greater than 10."))
}   else {
    print(paste0(n," is equal to 10"))
}
}
```

```{r}
g <- 0

while(g <= 10) {
  print(paste0(g, " is less than or equal to 10."))
  g <- g+1
}
```

#### Plotting and Figures

```{r}
gap <- read.csv("data/gapminder_all.csv")

gap
```

```{r}
plot(gap$gdpPercap_1997 ~ gap$pop_1997)
```

```{r}
#install.packages("ggplot2", dependencies = TRUE)

library(ggplot2)
```

```{r}
ggplot(data=gap, mapping=aes(y=gdpPercap_1997, x=pop_1997)) + geom_point()
```

```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/main/episodes/data/gapminder_data.csv")

gapminder
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(aes(color=continent))
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(aes(color=continent), alpha=0.25) + scale_x_log10() +
  geom_smooth(method=lm, color="black", fill="grey") + theme_classic()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=year, y=lifeExp, by=country)) + geom_line(mapping=aes(color=continent)) + geom_point() + theme_classic()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(color="darkblue", alpha=0.25, size=1) + scale_x_log10() +
  geom_smooth(method=lm, color="deeppink", fill="chartreuse", size=0.5) + theme_classic()
```

```{r}
ggplot(data=gapminder, mapping=aes(x=gdpPercap, y=lifeExp)) + geom_point(color="darkblue", alpha=0.25, size=1) +
  geom_smooth(method=loess, color="deeppink", fill="chartreuse", size=0.5) + theme_classic()
```

```{r}
unique(gapminder$continent)
```

```{r}
americas <- gapminder[gapminder$continent == "Americas", ]
```

```{r}
ggplot(data=americas, mapping = aes(x=year, y=lifeExp)) + 
  labs(
    x = "Year",
    y = "Life Expectancy (years)",
    title = "Life expectancy over time in the Americas"
  ) +
  geom_line(color="darkblue") + 
  facet_wrap( ~ country) +
  theme(axis.text.x = element_text(angle=90))
```

```{r}
ggsave("data/AmericasLifeExp.png")
```

```{r}
write.csv(americas, file ="data/americas.csv")
write.table(gapminder, file ="data/gapminder_web_20240930.csv", sep=",")
```

###### Fancy Plots

```{r}
#install.packages(c("ggridges", "viridis", "hrbrthemes"), dependencies=TRUE)
```

```{r}
library(ggridges)
library(viridis)
library(hrbrthemes)
```

```{r}
ggplot(lincoln_weather, aes(x = `Mean Temperature [F]`, y = `Month`, fill = after_stat(x))) +
  labs(title="Temperature in Lincoln, NE in 2016") +
  geom_density_ridges_gradient(scale = 3, rel_min_height=0.01) +
  scale_fill_viridis_c(name = "Temp. [F]", option = "C") +
  theme_ipsum()
```

```{r}
#install.packages(c("ggstatsplot", "tidyverse", "palmerpenguins"))
```

```{r}
library(tidyverse)
```

```{r}
data("penguins", package = "palmerpenguins")
penguins
```

```{r}
(penguins2 <- drop_na(penguins))
```

```{r}
ggplot(penguins2, aes(x=`bill_length_mm`, y=`bill_depth_mm`)) +
  geom_point() +
  geom_smooth(method="lm")
```

```{r}
ggplot(penguins2, aes(x=`bill_length_mm`, y=`bill_depth_mm`, color=`species`)) +
  geom_point() +
  geom_smooth(method="lm")
```

```{r}
library(ggstatsplot)

plt <- ggbetweenstats(
  data=penguins2,
  x=species,
  y=bill_length_mm
)

plt
```

```{r}
library(ggridges)
library(ggplot2)
library(viridis)

spider_data <- read.csv("data/spider_data_20221101.csv")
spider_data
```

```{r}
library(tidyverse)

spider_data %>% 
  arrange(year) %>%
  mutate(name=factor(family, levels=unique(family))) %>% 
  ggplot(aes(x = `year`, y = `family`, color = `family`, fill = `family`)) + 
  geom_density_ridges(alpha = 0.5, stat = "binline", bin = 25) +
  theme(
    legend.position = "none",
    strip.text.x = element_text(size = 5)
  )
```

```{r}
head(penguins2)
```

```{r}
penguins3 <- with(penguins2, cbind(bill_length_mm, bill_depth_mm, flipper_length_mm, body_mass_g))
```

```{r}
penguin_pca <- princomp(penguins3)
summary(penguin_pca)
```

```{r}
loadings(penguin_pca)
```

```{r}
biplot(penguin_pca)
```

```{r}
#install.packages("devtools")
library(devtools)

#install_github("vqv/ggbiplot")
library(ggbiplot)

ggbiplot(penguin_pca, 
         groups = penguins2$species, ellipse = TRUE, scale. = TRUE) +
  scale_color_discrete(name = "")
```

```{r}
# Get data:
#install.packages("gapminder")
#install.packages("gganimate")
#install.packages("gifski")
library(gapminder)
 
# Charge libraries:
library(ggplot2)
library(gganimate)
library(gifski)
 
# Make a ggplot, but add frame=year: one image per year
ggplot(gapminder, aes(gdpPercap, lifeExp, size = pop, colour = country)) +
  geom_point(alpha = 0.7, show.legend = FALSE) +
  scale_colour_manual(values = country_colors) +
  scale_size(range = c(2, 12)) +
  scale_x_log10() +
  facet_wrap(~continent) +
  # Here comes the gganimate specific bits
  labs(title = 'Year: {frame_time}', x = 'GDP per capita', y = 'life expectancy') +
  transition_time(year) +
  ease_aes('linear')
# Save at gif:
anim_save("data/271-ggplot2-animated-gif-chart-with-gganimate2.gif"
renderer = gifski_renderer())
```

### Statistics

```{r}
rand <- rnorm(50, 30, 8)
rand
```

```{r}
sum(rand)
```

```{r}
sum(rand)/length(rand)
```

```{r}
mean(rand)
```

```{r}
sort(rand)
```

```{r}
min(rand)
max(rand)
```

```{r}
cumsum(rand)
```

```{r}
diff(rand)
```

```{r}
sqrt(rand)
```

```{r}
log(rand)
```

```{r}
#things <- scan()
```

```{r}
things
```

#### Analyses

##### Univariate Statistics

1. Categorical Data
  - Barplots
  
```{r}
beer <- c(3,4,1,1,3,4,3,3,1,3,2,1,2,1,2,3,2,3,1,1,1,1,4,3,1)
length(beer)

table(beer)
```

```{r}
barplot(table(beer), xlab="Beer", ylab="Frequency", col="blue")
```

```{r}
barplot(table(beer)/length(beer), xlab="Beer", ylab="Frequency", col="blue")
```

```{r}
pie(table(beer), main = "Beer preference by students (n=25)")
```

2. Numerical data

```{r}
rand
stem(rand)
```

```{r}
stripchart(rand)
```

3. Measures of center

```{r}
mean(rand)
```

```{r}
median(rand)
```

```{r}
which(table(rand) == max(table(rand)))
```

4. Variation

  - Range
  
```{r}
range(rand)
```

  - Variance
  
```{r}
var(rand)
```

```{r}
sd(rand)
```

  -IQR

```{r}
IQR(rand)
```

  - z-scores
  
```{r}
scale(rand)
```

  - Summary
  
```{r}
summary(rand)
```

5. Plots 

  - Histograms
  
```{r}
hist(rand, breaks = 15, prob = TRUE)
```

  - Box plots
  
```{r}
boxplot(rand)
```

```{r}
#install.packages("vioplot")
library(vioplot)

vioplot(rand, col = "orange")
```

```{r}
vioplot(gapminder$lifeExp~gapminder$continent, col = "forestgreen")
```

###### Bivariate and multivariate statistics

  - Plotting and regression
    - Box plotting
    
```{r}
boxplot(gapminder$lifeExp~gapminder$continent, col = "sienna")
```


```{r}
vioplot(gapminder$lifeExp~gapminder$continent, col = "forestgreen")
```

```{r}
summary(rand)

t.test(rand, mu = 30)
```

```{r}
(spid.gen <- read.csv("data/spider_genitalia.csv", header = TRUE))
```

```{r}
vioplot(spid.gen$left.bulb/spid.gen$carapace.length~spid.gen$habitat)
```

```{r}
plot(spid.gen$left.bulb~spid.gen$right.bulb)
bulb.reg <- lm(spid.gen$left.bulb~spid.gen$right.bulb)
abline(bulb.reg)
```

```{r}
summary(bulb.reg)
```

```{r}
sqrt(0.5115)
```

```{r}
cor(spid.gen$left.bulb, spid.gen$right.bulb, method = "spearman")
```

```{r}
t.test(spid.gen$left.bulb/spid.gen$carapace.length~spid.gen$habitat)
```

```{r}
wilcox.test(spid.gen$left.bulb/spid.gen$carapace.length~spid.gen$habitat)
```

```{r}
ks.test(rand, "pnorm")
```

```{r}
shapiro.test(rand)
```

```{r}
qqnorm(rand, pch=16, col="red")
qqline(rand, lty=2, col="darkblue")
```

```{r}
summary(lm(spid.gen$left.bulb/spid.gen$carapace.length~spid.gen$habitat))
```

```{r}
summary(aov(spid.gen$left.bulb/spid.gen$carapace.length ~ spid.gen$habitat))
```

```{r}
habitat <- c("A", "A", "A", "A", "A", "B", "B", "B", "B", "B", "C", "C", "C", "C", "C")

parload <- c(10, 12, 8, 11, 14, 14, 16, 10, 9, 12, 18, 19, 20, 21, 22)

parhab <- cbind(habitat, parload)

parhab <- data.frame(parhab)

parhab$habitat <- as.factor(parhab$habitat)
parhab$parload <- as.numeric(parhab$parload)

parhab
```

```{r}
parhab.aov <- aov(parhab$parload ~ parhab$habitat)

summary(parhab.aov)
```

```{r}
(parhab.tukey <- TukeyHSD(parhab.aov))
```

```{r}
ggbetweenstats(parhab, x = habitat, y = parload)
```

```{r}
(pvals <- c(0.05, 0.005, 0.0005, 0.00005, 0.000005))

p.adjust(pvals, method = "bonferroni")

p.adjust(pvals, method = "fdr")
```

```{r}
gapminder.2007 <- gapminder[which(gapminder$year == 2007), ]

gapminder.2007.pop.gdpPercap.lifeExp.contintent <- lm(gapminder.2007$pop ~ gapminder.2007$gdpPercap + gapminder.2007$lifeExp + gapminder.2007$continent)

summary(gapminder.2007.pop.gdpPercap.lifeExp.contintent)
```

```{r}
#install.packages("MuMIn", dependencies = TRUE)
library(MuMIn)

# Example from Burnham and Anderson (2002), page 100:

# prevent fitting sub-models to different datasets
oop <- options(na.action = "na.fail")

dd <- dredge(gapminder.2007.pop.gdpPercap.lifeExp.contintent)
subset(dd, delta < 2)

dd
```


```{r}
# Visualize the model selection table:
  if(require(graphics)) {
    par(mar = c(3,4,6,4))
    plot(dd, labAsExpr = TRUE)
  }
```

```{r}
# Model average models with delta AICc < 2
model.avg(dd, subset = delta < 2)

#or as a 95% confidence set:
model.avg(dd, subset = cumsum(weight) <= .95) # get averaged coefficients

#'Best' model
summary(get.models(dd, 1)[[1]])
```

### Tidyverse

```{r}
library(tidyverse)
library(palmerpenguins)

head(penguins)
```

```{r}
weather <- read.csv("https://pasta.lternet.edu/package/data/eml/knb-lter-pal/28/8/375b34051b162d84516ec2d02f864675")
```

- Structure

```{r}
str(penguins)
```

```{r}
str(weather)
```

- Reducing Data to Set Dimensions

```{r}
colnames(penguins)
```

```{r}
Torg_2007_penguins <- penguins %>% select(species, island, year) %>%
  filter(year == 2007 & island == "Torgersen")

Torg_2007_penguins
```

```{r}
is.na(penguins$sex)
```

```{r}
summary(penguins %>% filter(!is.na(sex) & body_mass_g >= 5000))
```

```{r}
penguins_bill_mass_ratio <- penguins %>% mutate(bill.mass.ratio = bill_length_mm/body_mass_g)

vioplot(penguins_bill_mass_ratio$bill.mass.ratio)
```

```{r}
penguins %>% filter(!is.na(body_mass_g)) %>% 
  mutate(body_size_cat = case_when(
    body_mass_g >= 5000 ~ "big", 
    body_mass_g <= 3500 ~ "small",
    body_mass_g <5000 & body_mass_g >3500 ~ "medium",
    T ~ "ERROR"
))
```

```{r}
penguins %>% filter(!is.na(body_mass_g)) %>% 
  mutate(body_size_cat = case_when(
    body_mass_g >= 5000 ~ "big", 
    body_mass_g <= 3500 ~ "small",
    body_mass_g <5000 & body_mass_g >3500 ~ "medium",
    T ~ "ERROR"
  )) %>%
  filter(!is.na(sex)) %>%
  group_by(species, sex) %>%
  summarize(mean_body_mass_g = mean(body_mass_g, na.rm=T),
            sd_body_mass_g = sd(body_mass_g, na.rm=T))
```

- Manipulating Data

```{r}
str(weather)
```

```{r}
lubridate::yday("2024-10-28")
```

```{r}
weather %>% mutate(Date = ymd(Date))
```

```{r}
weather %>% mutate(Date = ymd(Date)) %>%
  mutate(Date = floor_date(Date, unit = 'year')) %>%
  group_by(Date) %>%
  summarize(mean(Temperature.Average..C.)) %>%
  
  ggplot(aes(x=Date, y=`mean(Temperature.Average..C.)`)) +
    geom_point(aes(color = year(Date))) +
    scale_color_viridis_c() +
    geom_line(color = "grey") +
    geom_smooth(method = "lm", linewidth = 1, linetype = 1, color = "red2", fill = "orange2")
```

- Joining Data Frames

```{r}
colnames(penguins)
```

```{r}
colnames(weather)
```

```{r}
(weather.sums <- weather %>%
  group_by(year(Date)) %>%
  summarize(avg.temp = mean(Temperature.Average..C., na.rm = T),
            max.temp = mean(Temperature.High..C., na.rm = T),
            min.temp = mean(Temperature.Low..C., na.rm = T)) %>%
  rename(year = `year(Date)`) 
)
```

```{r}
(penguins_weather.sums <- left_join(penguins, weather.sums, by = "year"))
```

```{r}
penguins_weather.sums %>%
  filter(!is.na(sex)) %>%
  
  ggplot(aes(x=sex, y=body_mass_g, color = sex, fill = sex)) +
    geom_violin() +
    scale_color_manual(values = c("orange3", "purple3")) +
    scale_fill_manual(values = c("orange", "purple")) +
    facet_wrap(~species + year) +
    labs(title = "Average penguin body size per species per year by sex",
         x="Sex",
         y ="Body Mass (g)") +
    theme_dark()
```

create a plot of average penguin body mass by annual min temp separated by species year and sex

```{r}
penguins_weather.sums %>%
  filter(!is.na(sex)) %>%
  group_by(species, sex, year) %>%
  summarize(mean.body.mass.g = mean(body_mass_g),
            mean.min.temp.c = mean(min.temp)) %>% 
  
  ggplot(aes(x=mean.min.temp.c, y=mean.body.mass.g, color = sex)) +
    geom_point() +
    scale_color_viridis_d() +
    geom_smooth(method = "lm", se=F) +
    facet_wrap(~species) +
    theme_classic() +
    labs(title="Average penguin body mass by average annual minimium temperature",
         x="Average minimum temperature (c)",
         y="Average body mass (g)")
```














